---
title: "Interactions"
author: "Taavi Päll and Ülo Maiväli"
date: "2021-10-02"
output: github_document
editor_options: 
  markdown: 
    wrap: 72
---

To model situation where the importance of one predictor depends upon another 
predictor we need interaction. 
Interaction allows parameters to be conditional on further aspects of the data. 
The simplest kind of interaction, a linear interaction, is built by extending 
the linear modeling strategy to parameters within the linear model.

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Loading some required libraries

```{r}
library(tidyverse)
library(here)
library(brms)
library(bayesplot)
library(tidybayes)
library(modelr)
library(rstan)
rstan_options(auto_write = TRUE)
options(mc.cores = parallel::detectCores())
```


## Data

To model interactions, we are going to use ruggedness data from **rethinking** 
package.

```{r}
(terrain_ruggedness <- read_csv(here("data/rugged.csv")))
```

The variable **rugged** is a terrain ruggedness index that quantifies the 
topographic heterogeneity of a landscape. 

The outcome variable here is the real gross domestic product per 
capita, from the year 2000, **rgdppc_2000**.

It's interesting that higher terrain ruggedness related to bad economies 
outside of Africa, but African economies may actually benefit from more rugged 
terrain.

We want to log transform GDP values and we need to create new variable to group 
countries locating in or out Africa.

We also rescale log GDP relative to average and ruggedness relative to maximum 
ruggedness.

To account whether nations are from Africa or not from Africa we create a new 
dummy variable "continent".

```{r}
ruggedness <- terrain_ruggedness %>% 
  mutate(
    log_gdp = log(rgdppc_2000),
    log_gdp = log_gdp / mean(log_gdp, na.rm = TRUE),
    rugged = rugged / max(rugged, na.rm = TRUE),
    continent = ifelse(cont_africa, "African nations", "Non-African nations")
    ) %>% 
  drop_na(log_gdp)
```

Let's make plots
```{r}
ruggedness %>% 
  ggplot(aes(rugged, log_gdp)) +
  geom_point() +
  geom_smooth(method = "lm") +
  facet_wrap(~continent) +
  labs(x = "Terrain ruggedness (std)", y = "log GDP (proportion of mean)")
```

We can see that in African nations is higher terrain ruggedness apparently 
related to higher GDP, whereas in Non-African nations trend is negative.

## Model

### Prior-only -- vague priors

Let's try to fit our first model and figure out priors, default look like this
```{r}
f <- log_gdp ~ rugged
get_prior(formula = f, data = ruggedness, family = gaussian())
```


Vague priors
```{r}
priors <- c(
  prior("normal(1, 1)", class = "Intercept"),
  prior("normal(0, 1)", class = "b"),
  prior("normal(0, 1)", class = "sigma")
)
mod1.0 <- brm(
  formula = log_gdp ~ rugged,
  data = ruggedness,
  family = gaussian(),
  prior = priors,
  chains = 3,
  file = here("models/log_gdp~rugged_prior_only1"),
  sample_prior = "only"
)
```

```{r}
plot(mod1.0)
```

```{r}
samples1.0 <- posterior_samples(mod1.0) %>% 
  sample_n(50)
```


```{r}
ruggedness %>% 
  ggplot(aes(rugged, log_gdp)) +
  geom_hline(yintercept = range(ruggedness$log_gdp), linetype = "dashed") +
  geom_abline(data = samples1.0, aes(slope = b_rugged, intercept = b_Intercept), size = 0.3, alpha = 0.3) +
  labs(x = "Ruggedness", y = "log GDP", subtitle = "intercept ~ normal(1, 1)\nb ~ normal(0, 1)") +
  theme_classic()
```

### Prior-only -- more stringent priors

Maximum possible change in log GDP is 0.6 within terrain ruggedness range in 
our data, but many regression lines display much more stronger effects, 
suggesting that we are using too lenient priors.

We can try narrower prior for b.

```{r}
priors <- c(
  prior("normal(1, 0.1)", class = "Intercept"),
  prior("normal(0, 0.3)", class = "b"),
  prior("normal(0, 1)", class = "sigma")
)
mod1.1 <- brm(
  formula = log_gdp ~ rugged,
  data = ruggedness,
  family = gaussian(),
  prior = priors,
  chains = 3,
  file = here("models/log_gdp~rugged_prior_only2"),
  sample_prior = "only"
)
```



```{r}
plot(mod1.1)
```

```{r}
samples1.1 <- posterior_samples(mod1.1) %>% 
  sample_n(50)
```

With stronger priors, picture looks much better.
```{r}
ruggedness %>% 
  ggplot(aes(rugged, log_gdp)) +
  geom_hline(yintercept = range(ruggedness$log_gdp), linetype = "dashed") +
  geom_abline(data = samples1.1, aes(slope = b_rugged, intercept = b_Intercept), size = 0.3, alpha = 0.3) +
  labs(x = "Ruggedness", y = "log GDP", subtitle = "intercept ~ normal(1, 0.1)\nb ~ normal(0, 0.3)") +
  theme_classic()
```

### One-predictor model with data

Model with data:
```{r}
mod1.2 <- brm(
  formula = log_gdp ~ rugged,
  data = ruggedness,
  family = gaussian(),
  prior = priors,
  chains = 3,
  file = here("models/log_gdp~rugged"),
  sample_prior = "yes"
)
```

```{r}
plot(mod1.2)
```



Posterior summary suggest that terrain ruggedness (b_rugged) has no effect on 
log_gdp.
```{r}
posterior_summary(mod1.2)
```

```{r}
samples1.2 <- posterior_samples(mod1.2) %>% 
  sample_n(50)
```


```{r}
ruggedness %>% 
  ggplot(aes(rugged, log_gdp)) +
  geom_hline(yintercept = range(ruggedness$log_gdp), linetype = "dashed") +
  geom_smooth(method = "lm", se = FALSE) +
  geom_abline(data = samples1.2, aes(slope = b_rugged, intercept = b_Intercept), size = 0.3, alpha = 0.3) +
  labs(x = "Ruggedness", y = "log GDP", subtitle = "intercept ~ normal(1, 0.1)\nb ~ normal(0, 0.3)") +
  theme_classic()
```

### Adding an another predictor

Let's see what happens if we add our dummy variable "continent" to model. 
We could also use original variable "cont_africa", but then we want to convert 
it to a category, otherwise it will be treatad as continuous, which it's not.
```{r}
mod1.3 <- update(
  mod1.2, 
  formula = log_gdp ~ rugged + continent, 
  newdata = ruggedness, 
  file = here("models/log_gdp~rugged+continent")
  )
```


```{r}
plot(mod1.3)
```


```{r}
summary(mod1.3)
```


```{r}
plot(conditional_effects(mod1.3), ask = FALSE, points = TRUE)
```


```{r}
samples1.3 <- ruggedness %>% 
  select(log_gdp, rugged, continent) %>% 
  add_epred_draws(mod1.3)
```

Including an indicator for African nations has no effect on the slope.
We can only see that African nations have lower log GDP.
```{r}
samples1.3 %>% 
  mean_qi() %>% 
  ggplot(aes(x = rugged)) +
  geom_point(aes(y = log_gdp, color = continent)) +
  geom_smooth(aes(y = .epred, color = continent), method = "lm", se = FALSE) +
  geom_ribbon(aes(ymin = .upper, ymax = .lower, fill = continent), alpha = 1/5) +
  theme_classic() +
  theme(legend.title = element_blank(), legend.position = "bottom")
```

### Plus interaction

To recover the change in slope that we saw in our first plot, we need to add a 
proper interaction effect. 

Let's see what happens when we add a term specifying an interaction between 
ruggedness and continent to our model:

```{r}
mod1.4 <- update(
  mod1.2, 
  formula = log_gdp ~ rugged + continent + rugged:continent, 
  newdata = ruggedness, 
  file = here("models/log_gdp~rugged+continent+rugged:continent")
  )
```


```{r}
plot(mod1.4)
```

```{r}
summary(mod1.4)
```

```{r}
samples1.4 <- ruggedness %>% 
  select(log_gdp, rugged, continent) %>% 
  add_epred_draws(mod1.4)
```

Now we can see from our model change in slope whether nation is from Africa or not.
```{r}
samples1.4 %>% 
  mean_qi() %>% 
  ggplot(aes(x = rugged)) +
  geom_point(aes(y = log_gdp)) +
  geom_smooth(aes(y = .epred)) +
  geom_ribbon(aes(ymin = .upper, ymax = .lower), alpha = 1/5) +
  facet_wrap(~continent, scales = "free_y")
```

We can also visualize difference in slopes using `conditional_effects()` 
function, by specifying that now we are interested only in interaction effect.
```{r}
plot(conditional_effects(mod1.4, effects = "rugged:continent"), 
     ask = FALSE, plot = FALSE, points = TRUE)[[1]] +
  theme(legend.title = element_blank(), legend.position = "bottom") +
  coord_fixed()
```


