---
title: "Logistic Regression -- Poisson models"
author: "Taavi Päll and Ülo Maiväli"
date: "2021-10-02"
output: github_document
editor_options: 
  markdown: 
    wrap: 72
---

Here we have a look how to fit Poisson models.

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Loading some required libraries

```{r}
library(tidyverse)
library(here)
library(brms)
library(bayesplot)
library(tidybayes)
library(modelr)
library(lubridate)
library(rstan)
rstan_options(auto_write = TRUE)
options(mc.cores = parallel::detectCores())
```


## Data

Estonian COVID-19 data on the daily number of new reported COVID-19 cases and deaths is from ECDC <https://www.ecdc.europa.eu/en/publications-data/data-daily-new-cases-covid-19-eueea-country>

```{r}
(covid_deaths_imported <- read_csv(here("data/covid19_deaths.csv")))
```

```{r}
(covid_deaths <- covid_deaths_imported %>% 
   mutate(
     dateRep = dmy(dateRep),
     time = yday(dateRep)
   )
 )
```



```{r}
covid_deaths %>% 
  ggplot() +
  geom_col(aes(dateRep, deaths)) +
  labs(x = "Date", y = "Number of COVID-19 deaths")
```

```{r}
covid_deaths %>% 
  ggplot() +
  geom_bar(aes(deaths))
```

```{r}
mean(covid_deaths$deaths)
```

```{r}
x <- rpois(1000, mean(covid_deaths$deaths))
qplot(x)
```


### Simulate counts first

Here we simulate some counts to demonstrate that in Poisson distribution 
the mean and the variance are equal.

```{r}
set.seed(11) 
tibble(y = rbinom(1e5, 1000, 1 / 1000)) %>% 
  summarise(
    y_mean = mean(y),
    y_variance = var(y)
    )
```

Poisson likelihood is 

$$y \sim Poisson(\lambda)$$

Since $\lambda$ is constrained to be positive, we typically use the log link. 
Thus the basic Poisson regression model would be

$$\text{deaths}_i \sim Poisson(\lambda_i)$$
$$log(\lambda_i) = \alpha + \beta \text{time}_i$$
$$\alpha \sim \text{Normal}(0, 10)$$
$$\beta \sim \text{Normal}(0, 1)$$

## Model

```{r}
f <- deaths ~ time
get_prior(formula = f, data = covid_deaths, family = poisson(link = "log")) # log link is default
```

```{r}
range(covid_deaths$deaths)
```

```{r}
table(covid_deaths$deaths)
```


```{r}
priors <- c(
  prior("normal(0, 0.2)", class = "Intercept"),
  prior("normal(0, 0.85)", class = "b")
)
mod1.0 <- brm(
  formula = deaths ~ time,
  data = covid_deaths,
  family = poisson(link = "log"),
  prior = priors,
  chains = 3,
  file = here("models/deaths~time_prior_only"),
  sample_prior = "only",
  file_refit = "on_change"
)
```


```{r}
summary(mod1.0)
```



```{r}
pp_check(mod1.0)
```

```{r}
y <- covid_deaths$deaths
yrep <- posterior_predict(mod1.0)
```


```{r}
ppc_dens_overlay(y, yrep[1:50, ])
```


Number of zeros.
```{r}
prop_zero <- function(x) mean(x == 0)
print(prop_zero(y))
```

```{r}
ppc_stat(y, yrep, stat = "prop_zero")
```

Mean and sd are correlated, which is common with count data, specially with Poisson distribution, where mean and variance are identical.
```{r}
ppc_stat_2d(y, yrep, stat = c("mean", "sd"))
```



```{r}
ppc_stat(y, yrep, stat = "mean", binwidth = 1)
```

```{r}
ppc_stat(y, yrep, stat = "max", binwidth = 1)
```


```{r}
covid_deaths %>% 
  add_epred_draws(mod1.0) %>% 
  ggplot() +
  geom_point(aes(deaths, .epred))
```



### Adding data

```{r}
mod1.1 <- brm(
  formula = deaths ~ time,
  data = covid_deaths,
  family = poisson(link = "log"),
  prior = priors,
  chains = 3,
  file = here("models/deaths~time"),
  sample_prior = "yes",
  file_refit = "on_change"
)
```


```{r}
plot(mod1.1)
```

```{r}
pp_check(mod1.1)
```


```{r}
plot(conditional_effects(mod1.1), plot = FALSE)[[1]] +
  geom_point(data = covid_deaths, aes(time, deaths), inherit.aes = FALSE)
```

### Splines

```{r}
mod1.2 <- update(mod1.1, formula = deaths ~ s(time, k = 6), file = here("models/deaths~s(time,k=6)"))
```

```{r}
plot(mod1.2)
```


```{r}
pp_check(mod1.2)
```


```{r}
samples1.2 <- covid_deaths %>% 
  add_epred_draws(mod1.2)
```


```{r}
samples1.2 %>% 
  ggplot() +
  stat_lineribbon(aes(time, .epred), alpha = 1/5) +
  geom_point(aes(time, deaths))
```

Same plot using conditional effects:
```{r}
plot(conditional_effects(mod1.2), ask = FALSE, points = TRUE)
```


Let's see what prediction looks like.
```{r}
# seq(min(covid_deaths$dateRep) - days(30), min(covid_deaths$dateRep) - 1, by = "days")


time <- c(covid_deaths$dateRep, seq(today(), today() + days(30), by = "days")) %>% 
  yday()
```


```{r}
tibble(time) %>% 
  add_predicted_draws(mod1.2) %>% 
    ggplot() +
  stat_lineribbon(aes(time, .prediction), alpha = 1/5) +
  geom_vline(xintercept = yday(today()), linetype = "dashed")
```


Here comes another approach from <http://avehtari.github.io/BDA_R_demos/demos_rstan/trafficdeaths.html>
```{r}
y_predict <- posterior_predict(mod1.2, newdata = tibble(time))
mu <- apply(t(y_predict), 1, quantile, c(0.05, 0.5, 0.95)) %>%
  t() %>% data.frame(x = time, .) %>% gather(pct, y, -x)
ggplot() +
  geom_point(aes(time, deaths), data = covid_deaths, size = 1) +
  geom_line(aes(x, y, linetype = pct), data = mu, color = 'red') +
  scale_linetype_manual(values = c(2,1,2)) +
  labs(x = 'Day', y = 'COVID-19 deaths') +
  guides(linetype = F)

```

