---
title: "Missing data"
author: "TP"
date: "15 10 2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


Imputing missing values is relatively straightforward with **brms** R package.
There are two possible approaches, 

(1) Impute missing values before the model fitting with multiple imputation, or 
(2) impute missing values on the fly during model fitting.


## Loading libraries

```{r}
library("tidyverse")
library("here")
library("brms")
library("mice")
library("naniar")
library("qgcomp")
library("rstan")
library("tidybayes")
rstan_options(auto_write = TRUE)
options(mc.cores = parallel::detectCores())
```


## Fitting model using pre-imputed data

Sensory nerve conductance (scv, m/s), distal latency time (sdl, ms) and 
amplitude (amp, mV) was measured from 2nd finger before and after five rounds of treatment 
with a certain wrist wrap.


```{r}
wrist <- read_csv(here("data/nervus_medianus_2nd_finger.csv")) %>% 
  rename(subj = id) %>% 
  mutate(tp = factor(tp, levels = c("pre", "post")))
```


Visualizing data.

```{r}
pj <- position_jitter(width = 0.1)
wrist %>%
  pivot_longer(c("sdl", "scv", "amp")) %>% 
  ggplot(aes(as.factor(tp), value)) +
  geom_boxplot() +
  geom_point(position = pj) +
  geom_line(aes(group = subj), alpha = 0.5) +
  facet_wrap(~ name)
```

Visualizing missing data points.

```{r}
vis_miss(wrist)
```


Missing data (NAs) in our case are from measurements below instrument limit of detection (LOD).
 
We can impute such missing readings using **mice** package and `leftcenslognorm` method (imported from **qgcomp**). 


```{r}
imputed <- mice(wrist, method = "leftcenslognorm", print = FALSE)
plot(imputed)
```


(have a look at the imputed dataset)


We can visualize imputed values relative to original data:
```{r}
comp <- complete(imputed, action = "long") %>% as_tibble()
comp %>% 
  pivot_longer(cols = c("sdl", "scv", "amp")) %>% 
  select(subj, tp, age, name, value) %>%
  distinct() %>% 
  group_by(subj, tp, age, name) %>% 
  add_count() %>% 
  mutate(Values = if_else(n > 1, "imputed", "original")) %>% 
  ggplot(aes(tp, value)) +
  geom_boxplot(outlier.shape = NA) +
  geom_point(aes(color = Values), position = pj) +
  geom_line(aes(group = subj), alpha = 0.5) +
  facet_wrap(~name, scales = "free_y") +
  labs(caption = "Lines denote individual trends.")
```



Our default priors, based on original data (get_prior cannot handle mice output).

```{r}
get_prior(amp ~ tp * age + (1| subj), data = wrist, family = gaussian())
```


Setting priors for our model.

```{r}
priors <- c(
  prior("normal(0, 0.5)", class = "b"),
  prior("normal(50, 10)", class = "Intercept"),
  prior("exponential(1)", class = "sigma")
)
```


Model fitting with multiple imputed data is straightforward with `brms_multiple` function:  

```{r}
fit <- brm_multiple(
  scv ~ tp * age + (1 | subj), 
  data = imputed, 
  family = gaussian(), 
  prior = priors, 
  chains = 2,
  file = here("models/scv ~ tp * age + (1 | subj)")
  )
```

Note that the model is fitted separately to each imputed data slice.

This is our model summary

```{r}
summary(fit)
```

Models fitted using multiple imputed data, can show Rhats higher than 1.1, which might be false positive. 
In our case Rhats seem reasonable anyway.

```{r}
plot(fit, pars = "^b")
```

> Note that model was fitted to each imputed dataset (number of chains!).
 
We can see that our interaction-model suggest some improvement in sensory nerve conductance after treatment.

```{r}
plot(conditional_effects(fit, effects = "age:tp"), plot= FALSE)[[1]] +
  labs(x = "Patient age, years", y = "Sensory nerve conductance, m/s")
```


> The advantage of multiple imputation is that one can apply it to all kinds of models, since model fitting functions do not need to know that the data sets were imputed. 
The drawback is the extra amount of time required for model fitting. 

## Impute missing values on the fly during model fitting

Our data is from Estonian Sars-Cov-2 sequencing project.

We have variables for sample collection date, county of residence, age/age groups, vaccine brand, vaccination status (at least one shot), qPCR virus load proxy (orf1ab_ct) where smaller numbers mean higher virus load, presence of symptoms and assigned virus lineage where "None" usually means that obtained sequence was not good enough to assign a lineage (e.g. sequence quality is also related to virus load, less virus means that there might be not enough sequencing reads to assemble a virus genome to be used in lineage assignment).


```{r}
koro <- read_csv(here("data/korogenoest.csv"))
```

Ok, this dataset is not that bad and is filtered to include at least collection date for each sample (minimum amount of metadata to upload sequence to ENA).

```{r}
vis_miss(koro)
```



```{r}
data <- koro %>% 
  drop_na(orf1ab_ct, lineage, symptomatic, age)
data %>% 
  vis_miss()
```


```{r}
get_prior(orf1ab_ct ~ lineage + symptomatic + age, data = data, family = gaussian())
```

First, let's fit our model using no-na dataset:

```{r}
fit0 <- brm(orf1ab_ct ~ lineage + symptomatic + age, 
            data = data, 
            family = gaussian(), 
            prior = prior("normal(0, 2.5)", class = "b"),
            file = here("models/orf1ab_ct ~ lineage + symptomatic + age"),
            chains = 2)
```

Model summary of no-na model

```{r}
summary(fit0)
```



```{r, fig.height=7}
plot(conditional_effects(fit0, effects = "lineage"), plot = FALSE)[[1]] +
  coord_flip() +
  labs(x = "Covid-19 lineage")
```

We can see that "None" shows highest Ct values indicating very low virus titer.
Highest Ct values apparently come from individuals infected with AY.9 and AY.4 variants (a.k.a Delta/India)

Using age10 classes may even result in a better fit, but then we don't have it as continuous variable, which is a drawback in following imputation business.


### Model with imputed data

Let's try to fit another model with "complete" dataset and impute missing values during model fitting.

Nevertheless, we need to drop observations with 
- missing lineage, because we cannot impute them with mi (Error: Argument 'mi' is not supported for family 'categorical(logit)').
- Likewise, we have a problem with predictor symptomatic "Error: Argument 'mi' is not supported for family 'bernoulli(logit)'."

Model specification for on-the-fly imputation is similar to multivariate model specification:
```{r}
data1 <- koro %>% 
  drop_na(lineage, symptomatic)
ct <- bf(orf1ab_ct | mi() ~ lineage + symptomatic + mi(age))
age <- bf(age | mi() ~ lineage + symptomatic)
fit1 <- brm(formula = ct + age + set_rescor(FALSE), 
            data = data1, 
            family = gaussian(), 
            prior = prior("normal(0, 2.5)", class = "b"),
            file = here("models/bf(orf1ab_ct | mi() ~ lineage + symptomatic + mi(age)) + bf(age | mi() ~ lineage + symptomatic)"),
            chains = 2)
```

Now model summary shows that we have coefficients for both response variables.

```{r}
summary(fit1)
```


```{r, fig.height=7}
plot(conditional_effects(fit1, effects = "lineage"), plot = FALSE)[[1]] +
  coord_flip() +
  labs(x = "Covid-19 lineage", title = "Imputed model")
```

Let's compare these two models side-by-side

```{r}
# draws from the initial fit
f0d <- as.data.frame(fit0) %>% 
  select(b_Intercept, starts_with("b_lineage")) %>% 
  pivot_longer(cols = starts_with("b_lineage")) %>% 
  mutate(ct = b_Intercept + value,
         lineage = str_remove(name, "b_lineage"))
# draws from imputation-fit
f1d <- as.data.frame(fit1) %>% 
  select(b_orf1abct_Intercept, starts_with("b_orf1abct_lineage")) %>% # here we need to use orf1abct draws
  pivot_longer(cols = starts_with("b_orf1abct_lineage")) %>% 
  mutate(ct = b_orf1abct_Intercept + value,
         lineage = str_remove(name, "b_orf1abct_lineage"))
```

We can see that in this case the models are very similar (we had only few missing values after all)

```{r, fig.height=10}
bind_rows(f0d, f1d, .id = "model") %>% 
  mutate(model = if_else(model == 1, "drop_na", "imputation")) %>% 
  ggplot() +
  stat_pointinterval(aes(lineage, ct, color = model), position = "dodge") +
  coord_flip() +
  labs(x = "Covid-19 lineage", caption = "Points denote median and\nbold and thin lines 67% and 95% qi, respectively.")
```


## Multivariate models

As imputation during model fitting was similar to multivariate model fitting, we can easily fit our first model also as a simple multivariate model.

We had three simultaneously measured parameters -- sdl, amp and scv and we can model them together.
Let's reduce complexity and keep only subjects with complete data.

```{r}
wdata <- wrist %>% 
  drop_na() %>% 
  add_count(subj) %>% 
  filter(n == 2)
```


```{r}
vis_miss(wdata)
```



```{r}
fit <- brm(
  bf(mvbind(scv, amp, sdl) ~ tp * age + (1 | subj)) + set_rescor(FALSE), 
  data = wdata, 
  family = gaussian(), 
  chains = 2,
  file = here("models/mvbind(scv, amp, sdl) ~ tp * age + (1 | subj)")
  )
```

We can see that each of the parameters now have the corresponding response variable as prefix.

```{r}
summary(fit)
```


Let’s take a look at some posterior-predictive checks, which give us a first impression of the model fit.

```{r}
pp_check(fit, resp = "scv")
```


```{r}
pp_check(fit, resp = "amp")
```


```{r}
pp_check(fit, resp = "sdl")
```

Amp and scv look good, but not so much sdl:

```{r}
bayes_R2(fit)
```

One can specify different response distribution family for different responses: 

```{r}
scv <- bf(scv ~ tp * age + (1 | subj))
amp <- bf(amp ~ tp * age + (1 | subj))
sdl <- bf(sdl ~ tp * age + (1 | subj)) + skew_normal()
fit3 <- brm(
  scv + amp + sdl + set_rescor(FALSE), 
  data = wdata, 
  family = gaussian(), 
  chains = 2,
  file = here("models/scv + amp + sdl + set_rescor(FALSE)")
  )
```

```{r}
pp_check(fit3, resp = "sdl")
```

Seems little better.

```{r}
plot(conditional_effects(fit3, "age:tp", resp = "amp"))
```

Based on amp, we can imagine that 40+ patients benefit more from the therapy (amp increase is a positive dynamics)

```{r}
plot(conditional_effects(fit3, "age:tp", resp = "scv"))
```

Nerve conductance speed might increase in all age groups in response to therapy.

```{r}
plot(conditional_effects(fit3, "age:tp", resp = "sdl"))
```


sdl decreases when nerve conductance improves.
Together, all these parameters suggest a positive response to the treatment.

