---
title: "Multilevel ANOVA"
output:
  html_document:
    df_print: paged
---

```{r}
library(tidyverse)
library(here)
library(brms)
library(tidybayes)
options(mc.cores = parallel::detectCores())
```

## Multilevel modeling

Model:

y - pupils score
sch - school ID

$y_i \sim N(\mu, \sigma)$ - generates y-values from a normal distribution of scores. What if sigma = 0? Then all y-s are identical = mu. What if sigma = inf? Then mu loses meaning and y comes from runif(-Inf, Inf).

Now, what about different schools? Each school has its mean(score) and these mean scores have some sort of variation, which means that we can calculate their SD. 


Moreover, we can build meta-distributions of schools means and SD-s like this 

$$N(\mu (mean(score)), \sigma (mean(score))$$


$$N(\mu (sd(score)), \sigma (sd(score)))$$

Here we assume that schools come from a normal distribution of school means/sd-s. If the meta-sigma for school means = 0, then all schools are identical and we can ignore the sch variable in regression. If this meta-sigma = Inf, then all schools are completely separate and there is no information about sch A in sch B and vice versa. 

$$y_i \sim N(\alpha, \sigma_y)$$



Then we model each sch. independently of each other. With all other sigma_sch values we want to model scores and mean scores together so that schools could exchange information during model fitting: within school j 

$$y_i \sim N(\mu_j, \sigma_y) ~ or ~ y \sim 0 + sch$$ in r model language.

or 

$$y_i \sim N(\mu_j, \sigma_{y[j]}) ~ or ~ bf(y \sim 0 + sch, sigma \sim 0 + sch)$$


+ Multilevel models assume exhangeable non-ordinal grouping structures in the data (groups come as a random sample from a population of groups). This meta-population of groups is also modelled on data.

+ they assume more than one source of variation

+ they assume intra-group correlated data (not independent data)

+ they do not assume homoscedacity (SD-s can vary between groups)

+ they are interaction models

+ they model dependencies between the levels in the model

+ they partially pool information between groups, which leads to shrinkage of coefs. (and biased estimates)

+ linearity & normality assumptions apply to every level of the model.

+ multilevel models divide variation between intra-group and inter-group variation

+ they model the variation of individual group level regression coeficients (intercepts and means) 

+ they allow to more precisely estimate these regression coefs themselves, especially if N differs from group to group.

## Start by a simulation

```{r}
gr <- c(rep("A", times = 1), 
        rep("B", times = 2), 
        rep("C", times = 3), 
        rep("D", times = 4), 
        rep("E", times = 5), 
        rep("F", times = 6), 
        rep("G", times = 7), 
        rep("H", times = 8),
        rep("I", times = 9),
        rep("J", times = 10),
        rep("exp", times = 8))

set.seed(1)
value <- c(rlnorm(55), rlnorm(8, 1.5))

data1 <- data.frame(gr, value)

m_sim1 <- brm(value ~ 0 + gr, data = data1, family = lognormal(), file = here("models/m_sim1"))
m_sim2 <- brm(value ~ 0 + (1 | gr), data = data1, family = lognormal(), file = here("models/m_sim2"))
m_sim3 <- brm(bf(value ~ 0 + (1 | gr), sigma ~ 0 + (1 | gr)), 
              data = data1, family = lognormal(), file = here("models/m_sim3"))
m_sim4 <- brm(bf(value ~ 0 + (1 | id | gr), sigma ~ 0 + (1 | id | gr)), 
              data = data1, family = lognormal(), file = here("models/m_sim4"))
```


```{r}
posterior_summary(m_sim1)
```


```{r}
posterior_summary(m_sim2)
```

```{r}
posterior_summary(m_sim3)
```


```{r}
posterior_summary(m_sim4)
```


```{r}
m_sim2p <- posterior_summary(m_sim2) %>% 
  exp() %>% 
  as.data.frame() %>% 
  rownames_to_column() %>% 
  filter(str_detect(rowname, "r_gr") ) 
m_sim2p$rowname <- m_sim2p$rowname %>% 
  str_replace_all("r_gr\\[", "")
m_sim2p$rowname <- m_sim2p$rowname %>% 
  str_replace_all(",Intercept\\]", "")
m_sim2p$index <- "m_sim2"

m_sim1p <- posterior_summary(m_sim1) %>% 
  as.data.frame()%>% 
  exp() %>% 
  rownames_to_column() %>% 
  filter(str_detect(rowname, "b_gr") )
m_sim1p$rowname <- m_sim1p$rowname %>% 
  str_replace_all("b_gr", "")
m_sim1p$index <- "m_sim1"

m_sim <- bind_rows(m_sim1p, m_sim2p)
```


```{r}
ggplot(m_sim, aes(rowname, Estimate, color=index)) + 
  geom_pointinterval(aes(ymin=`Q2.5`, ymax=`Q97.5`), position = "dodge") + 
  xlab(NULL)+ ylab("median PT hrs") + 
  coord_flip()
```



## Now to HF data


```{r}
HF_data <- read_delim(here("data/HF_data.csv"), 
                      delim = ";", 
                      escape_double = FALSE, 
                      trim_ws = TRUE) %>% 
  sample_frac(0.05)
HF_data <- HF_data %>% 
  filter(county != "homeless")
HF_data <- HF_data %>% 
  select(postacute_therapy, dementia, county, postacute_therapy_binned, age, sex, age_binned10, postacute_LOS, postacute_hospitalisations)
HF_data$dementia <- as.factor(HF_data$dementia)
HF_data
```

m1. intercept-only model for proportion who get therapy
m2 single-level, county is regressor
m3 - 2-level anova-like model


### ANOVA for proportion of getting PT

```{r}
m1 <- brm(
  data = HF_data, 
  formula = postacute_therapy_binned ~ 1, 
  family = bernoulli(), 
  chains = 2,
  file = here("models/m1"))
m1l <- loo(m1)

m2 <- brm(
  data = HF_data, 
  formula = postacute_therapy_binned ~ 0 + county, 
  family = bernoulli(), 
  chains = 2, 
  file = here("models/m2"))
m2l <- loo(m2)

m3 <- brm(data = HF_data, postacute_therapy_binned ~ 0 + (1 | county), 
          family = bernoulli(), 
          chains = 2, 
          file = here("models/m3"))
m3l <- loo(m3)
```

```{r}
pp_check(m1)
```

```{r}
pp_check(m2)
```


```{r}
pp_check(m3)
```


```{r}
loo_compare(m1l, m2l, m3l)
```

```{r}
m3ps <- posterior_summary(m3) %>% inv_logit_scaled() %>% as.data.frame() %>% rownames_to_column() %>% filter(str_detect(rowname, "r_county") ) 
m3ps$rowname <- m3ps$rowname %>%  str_replace_all("r_county\\[", "")
m3ps$rowname <- m3ps$rowname %>%  str_replace_all(",Intercept\\]", "")
m3ps$index <- "m3"

m2ps <- posterior_summary(m2) %>% as.data.frame()%>% inv_logit_scaled() %>% rownames_to_column() %>% filter(str_detect(rowname, "b_county") )
m2ps$rowname <- m2ps$rowname %>%  str_replace_all("b_county", "")
m2ps$index <- "m2"

m23 <- bind_rows(m3ps, m2ps)

ggplot(m23, aes(rowname, Estimate, color=index)) + geom_pointinterval(aes(ymin=`Q2.5`, ymax=`Q97.5`), position = "dodge")+ xlab(NULL)+ ylab("Pr of getting PT")+ coord_flip()
```

```{r}
conditional_effects(m2)
```


### ANOVA for PT hrs

```{r}
(m_anova_pt <- brm(
  data = HF_data %>% filter(postacute_therapy>0),
  formula = postacute_therapy~0 + (1|county), 
  family = lognormal(), 
  chains = 2,
  file = here("models/m_anova_pt")))
```

note that 2nd level SD > data level SD! This means that there is more variation between mean PT hrs of different counties than there is between the patients in individual counties.


```{r}
m_anova_pts <- posterior_summary(m_anova_pt) %>% exp() %>% as.data.frame() %>% rownames_to_column() %>% filter(str_detect(rowname, "r_county") ) 
m_anova_pts$rowname <- m_anova_pts$rowname %>%  str_replace_all("r_county\\[", "")
m_anova_pts$rowname <- m_anova_pts$rowname %>%  str_replace_all(",Intercept\\]", "")

d <- HF_data %>% filter(postacute_therapy>0) %>% 
  group_by(county) %>% 
  summarise(Estimate= exp(mean(log(postacute_therapy)))) %>% 
  #geometric mean
  rename(rowname= county)

ggplot() + 
  geom_pointinterval(data = m_anova_pts, aes(rowname, Estimate, ymin=`Q2.5`, ymax=`Q97.5`), size=3)+
  geom_point(data=d, aes(rowname, Estimate), size=0.9, color="red")

```

Here we plot geometric means.

